\chapter{In praise of inequalities}
\label{chapter20}
\begin{theorem}
  \label{ch20theoremI}
  \lean{cauchy_schwarz_inequality}
  \leanok
Let $\langle a, b \rangle$ be an inner product on a real vector space $V$ (with the norm $|a|^2 := \langle a, a \rangle$). Then
\[
\langle a, b \rangle^2 \leq |a|^2 |b|^2
\]
holds for all vectors $a, b \in V$, with equality if and only if $a$ and $b$ are linearly dependent.
\end{theorem}
\begin{proof}
  \leanok
The following (folklore) proof is probably the shortest. Consider the quadratic function
\[
|x a + b|^2 = x^2 |a|^2 + 2x \langle a, b \rangle + |b|^2
\]
in the variable $x$. We may assume $a \neq 0$. If $b = \lambda a$, then clearly
\[
\langle a, b \rangle^2 = |a|^2 |b|^2.
\]
If, on the other hand, $a$ and $b$ are linearly independent, then $|x a + b|^2 > 0$ for all $x$,
and thus the discriminant $\langle a, b \rangle^2 - |a|^2 |b|^2$ is less than 0.
\end{proof}

\begin{theorem}[First proof]
  \label{ch20theoremIIproof1}
  \lean{harmonic_geometric_arithmetic₁}
  \leanok
  Let $a_1, \dots, a_n$ be positive real numbers, then
  \[
  \frac{n}{\frac{1}{a_1}+\dots+\frac{1}{a_n}} \le
    \sqrt[n]{a_1a_2\dots a_n} \le
    \frac{a_1+\dots+ a_n}{n}
  \]
 with equality in both cases if and only if all $a_i$'s are equal.
\end{theorem}
 \begin{proof}
  \leanok
  The following beautiful nonstandard induction proof is attributed to Cauchy.
  Let $P(n)$ be the statement of the second inequality, written in the form
  \[
  a_1 a_2 \cdots a_n \le \left(\frac{a_1 + \cdots + a_n}{n}\right)^n.
  \]
  For $n=2$, we have $a_1 a_2 \le \left(\frac{a_1+a_2}{2}\right)^2 \iff (a_1-a_2)^2 \ge 0$, which is true.
  Now we proceed in the following two steps:
  \begin{enumerate}
    \item[(A)] $P(n) \implies P(n-1)$
    \item[(B)] $P(n)$ and $P(2) \implies P(2n)$
  \end{enumerate}
  which will clearly imply the full result.

  To prove (A), set $A := \frac{\sum_{k=1}^{n-1} a_k}{n-1}$, then
  \[
  \left(\prod_{k=1}^{n-1} a_k\right) A \stackrel{P(n)}{\le} \left(\frac{\sum_{k=1}^{n-1} a_k + A}{n}\right)^n = \left(\frac{(n-1)A + A}{n}\right)^n = A^n
  \]
  and hence $\prod_{k=1}^{n-1} a_k \le A^{n-1}$.

  For (B), we see
  \[
  \prod_{k=1}^{2n} a_k = \left(\prod_{k=1}^{n} a_k\right)\left(\prod_{k=n+1}^{2n} a_k\right)
  \stackrel{P(n)}{\le} \left(\frac{\sum_{k=1}^{n} a_k}{n}\right)^n \left(\frac{\sum_{k=n+1}^{2n} a_k}{n}\right)^n
  \stackrel{P(2)}{\le} \left(\frac{\sum_{k=1}^{2n} a_k}{2n}\right)^{2n}.
  \]
  The condition for equality is derived just as easily.

  The left-hand inequality, between the harmonic and the geometric mean,
  follows now by considering $\frac{1}{a_1}, \dots, \frac{1}{a_n}$.
\end{proof}

\begin{theorem}[Another Proof]
  \label{ch20theoremIIproof2}
  \lean{harmonic_geometric_arithmetic₂}
  \leanok
  Let $a_1, \dots, a_n$ be positive real numbers, then
  \[
  \frac{n}{\frac{1}{a_1}+\dots+\frac{1}{a_n}} \le
    \sqrt[n]{a_1a_2\dots a_n} \le
    \frac{a_1+\dots+ a_n}{n}
  \]
 with equality in both cases if and only if all $a_i$'s are equal.
\end{theorem}
 \begin{proof}
  \leanok
  Of the many other proofs of the arithmetic-geometric mean inequality, let us single out
  a particularly striking one by Horst Alzer, with some shortenings due to France Dacar.
  As a matter of fact, this proof yields the stronger inequality
  \[
  a_1^{p_1} a_2^{p_2} \cdots a_n^{p_n} \le p_1 a_1 + p_2 a_2 + \cdots + p_n a_n
  \]
  for any positive numbers $a_1, \dots, a_n$, $p_1, \dots, p_n$ with $\sum_{i=1}^n p_i = 1$.
  Let us denote the expression on the left side by $G$, and on the right side by $A$.
  Fix $c > 0$ and define the function $f(t) := \frac{1}{c} - \frac{1}{t}$ on $\mathbb{R}_{>0}$.
  Since $f(t) < 0$ for $t < c$ and $f(t) > 0$ for $t > c$, we get the inequality
  \[
  \int_c^x f(t)\,dt \ge 0
  \]
  for every $x > 0$, with equality if and only if $x = c$.
  Now
  \[
  0 \le \int_c^x f(t)\,dt = \left[\frac{t}{c} - \log t\right]_c^x = \frac{x}{c} - 1 - \log\frac{x}{c},
  \]
  and setting $c = G$ and $x = a_i$ we conclude that
  \[
  \frac{a_i}{G} - 1 \ge \log a_i - \log G \qquad \text{for } i = 1, 2, \dots, n.
  \]
  Multiplying this inequality by $p_i$ and summing over all $i$ gives
  \[
  \sum_{i=1}^n p_i \frac{a_i}{G} - \sum_{i=1}^n p_i \ge \sum_{i=1}^n p_i \log a_i - \sum_{i=1}^n p_i \log G.
  \]
  With $\sum p_i = 1$, the left side equals $\frac{A}{G} - 1$, while the right side is
  $\log\left(\prod a_i^{p_i}\right) - \log G = \log G - \log G = 0$.
  We conclude $\frac{A}{G} - 1 \ge 0$, which is $A \ge G$.
  In the case of equality, all inequalities must be equalities, which implies $a_1 = \cdots = a_n = G$.
\end{proof}

\begin{theorem}[Still another Proof]
  \label{ch20theoremIIproof3}
  \lean{harmonic_geometric_arithmetic₃}
  \leanok
  Let $a_1, \dots, a_n$ be positive real numbers, then
  \[
  \frac{n}{\frac{1}{a_1}+\dots+\frac{1}{a_n}} \le
    \sqrt[n]{a_1a_2\dots a_n} \le
    \frac{a_1+\dots+ a_n}{n}
  \]
 with equality in both cases if and only if all $a_i$'s are equal.
\end{theorem}
 \begin{proof}
  \leanok
  There is another nice proof, due to Michael D.\ Hirschhorn. It uses Bernoulli's inequality, which says
  \[
  (1+t)^{n+1} \ge 1 + (n+1)t \qquad \text{for real } t \ge -1.
  \]
  Suppose $a_1, a_2, \dots, a_{n+1} > 0$ and set
  \[
  t = \frac{\frac{a_1 + \cdots + a_{n+1}}{n+1}}{\frac{a_1 + \cdots + a_n}{n}} - 1.
  \]
  By Bernoulli,
  \[
  \left(\frac{\frac{a_1 + \cdots + a_{n+1}}{n+1}}{\frac{a_1 + \cdots + a_n}{n}}\right)^{n+1}
  \ge 1 + (n+1)\left(\frac{\frac{a_1 + \cdots + a_{n+1}}{n+1}}{\frac{a_1 + \cdots + a_n}{n}} - 1\right)
  = \frac{n\, a_{n+1}}{a_1 + \cdots + a_n},
  \]
  which translates into
  \[
  \left(\frac{a_1 + \cdots + a_{n+1}}{n+1}\right)^{n+1} \ge a_{n+1} \left(\frac{a_1 + \cdots + a_n}{n}\right)^n,
  \]
  and the arithmetic-geometric mean inequality follows by induction.
\end{proof}

\begin{theorem}
  \label{ch20theorem1}
  \lean{laguerre_root_bound}
  \leanok
  Suppose all roots of the polynomial
  $x^n + a_{n-1}x^{n-1} + \dots + a_0$ are real.
  Then the roots are contained in the interval with the endpoints
  \[
  -\frac{a_{n-1}}{n}\pm\frac{n-1}{n}\sqrt{a_{n-1}^2 - \frac{2n}{n-1}a_{n-2}}.
  \]
\end{theorem}
\begin{proof}
  \leanok
  Let $y$ be one of the roots and $y_1, \dots, y_{n-1}$ the others. Then
  the polynomial is $(x-y)(x-y_1)\cdots(x-y_{n-1})$. Thus by comparing coefficients
  \begin{align*}
  -a_{n-1} &= y + y_1 + \cdots + y_{n-1}, \\
  a_{n-2} &= y(y_1 + \cdots + y_{n-1}) + \sum_{i<j} y_i y_j,
  \end{align*}
  and so $a_{n-1}^2 - 2a_{n-2} - y^2 = \sum_{i=1}^{n-1} y_i^2$.

  By Cauchy's inequality applied to $(y_1, \dots, y_{n-1})$ and $(1, \dots, 1)$,
  \[
  (a_{n-1} + y)^2 = (y_1 + \cdots + y_{n-1})^2 \le (n-1)\sum_{i=1}^{n-1} y_i^2 = (n-1)(a_{n-1}^2 - 2a_{n-2} - y^2),
  \]
  or
  \[
  y^2 + \frac{2a_{n-1}}{n}y + \frac{2(n-1)}{n}a_{n-2} - \frac{n-2}{n}a_{n-1}^2 \le 0.
  \]
  Thus $y$ (and hence all $y_i$) lie between the two roots of the quadratic function,
  and these roots are our bounds.

  Note: the Lean formalization proves the key Cauchy--Schwarz step yielding the quadratic inequality
  $n \cdot y^2 - 2S \cdot y + S^2 \le (n-1) \cdot Q$ where $S = \sum y_i$ and $Q = \sum y_i^2$,
  from which the endpoint formula follows by solving the quadratic.
\end{proof}



\begin{theorem}
  \label{ch20theorem2}
  \lean{erdos_gallai_A_ge_two_thirds_T}
  \leanok
  Let $f(x)$ be a real polynomial of degree $n \ge 2$ with only real
  roots, such that $f(x)> 0$ for $ -1 < x < 1$ amd $f(-1) = f(1) = 0$. Then
  \[
  \frac{2}{3}T \le A \le \frac{2}{3}R,
  \]
  and equality holds in both cases only for $n=2$.
\end{theorem}
\begin{proof}
  \uses{ch20theorem2}
  \leanok
  \textbf{Proof of $\frac{2}{3}T \le A$.}
  Since $f(x)$ has only real roots, and none of them in the open interval $(-1,1)$, it can be written
  (apart from a constant positive factor which cancels out) in the form
  \[
  f(x) = (1-x^2)\prod_i (\alpha_i - x)\prod_j (\beta_j + x)
  \]
  with $\alpha_i \ge 1$, $\beta_j \ge 1$. Hence $A = \int_{-1}^{1} f(x)\,dx$.

  By making the substitution $x \to -x$, we find that also
  $A = \int_{-1}^{1} (1-x^2)\prod_i(\alpha_i + x)\prod_j(\beta_j - x)\,dx$,
  and hence by the inequality of the arithmetic and the geometric mean
  (note that all factors are $\ge 0$):
  \begin{align*}
  A &= \int_{-1}^{1} \frac{1}{2}\left[(1-x^2)\prod(\alpha_i - x)\prod(\beta_j + x) + (1-x^2)\prod(\alpha_i + x)\prod(\beta_j - x)\right] dx \\
  &\ge \int_{-1}^{1} (1-x^2)\left[\prod(\alpha_i^2 - x^2)\prod(\beta_j^2 - x^2)\right]^{1/2} dx \\
  &\ge \int_{-1}^{1} (1-x^2)\left[\prod(\alpha_i^2 - 1)\prod(\beta_j^2 - 1)\right]^{1/2} dx \\
  &= \frac{4}{3}\left[\prod(\alpha_i^2 - 1)\prod(\beta_j^2 - 1)\right]^{1/2}.
  \end{align*}

  Computing $f'(1)$ and $f'(-1)$ from the product form, we find
  $f'(1) = -2\prod(\alpha_i - 1)\prod(\beta_j + 1)$ and
  $f'(-1) = 2\prod(\alpha_i + 1)\prod(\beta_j - 1)$,
  hence $-f'(1)f'(-1) = 4\prod(\alpha_i^2-1)\prod(\beta_j^2-1)$, so
  \[
  A \ge \frac{2}{3}(-f'(1)f'(-1))^{1/2}.
  \]
  Applying the inequality of the harmonic and geometric mean to $-f'(1)$ and $f'(-1)$, we arrive by
  formula~(2) for $T$ at
  \[
  A \ge \frac{2}{3}\cdot\frac{2}{\frac{1}{-f'(1)} + \frac{1}{f'(-1)}} = \frac{4}{3}\cdot\frac{f'(1)f'(-1)}{f'(1)-f'(-1)} = \frac{2}{3}T.
  \]

  \textbf{The inequality $A \le \frac{2}{3}R$:} The book notes ``The reader is invited to search for
  an equally inspired proof of the second inequality in Theorem~2.'' This half is not formalized in Lean.
\end{proof}

\begin{theorem}
  \label{ch20theorem3proof1}
  \lean{mantel}
  Suppose $G$ is a graph on $n$ vertices without triangles. Then $G$
  has at most $\frac{n^2}{4}$ edges, and equality holds only when $n$ is even and $G$ is the
  complete bipartite graph $K_{n/2, n/2}$.
\end{theorem}
\begin{proof}
  This proof, using Cauchy's inequality, is due to Mantel.
  Let $V = \{1, \dots, n\}$ be the vertex set and $E$ the edge set of $G$.
  By $d_i$ we denote the degree of $i$, hence $\sum_{i \in V} d_i = 2|E|$
  (see chapter \ref{chapter28}). Suppose $ij$ is an edge.
  Since $G$ has no triangles, we find $d_i + d_j \leq n$ since no vertex is a
  neighbor of both $i$ and $j$.

  It follows that
  \[
  \sum_{ij \in E} (d_i + d_j) \leq n|E|.
  \]
  Note that $d_i$ appears exactly $d_i$ times in the sum, so we get
  \[
  n|E| \geq \sum_{ij \in E} (d_i + d_j) = \sum_{i \in V} d_i^2,
  \]
  and hence with Cauchy's inequality applied to the vectors $(d_1, \dots, d_n)$ and $(1, \dots, 1)$,
  \[
  n|E| \geq \sum_{i \in V} d_i^2 \geq \frac{\left( \sum d_i \right)^2}{n} = \frac{4|E|^2}{n},
  \]
  and the result follows. In the case of equality we find $d_i = d_j$ for all $i, j$, and further
  $d_i = \frac{n}{2}$ (since $d_i + d_j = n$). Since $G$ is triangle-free, $G = K_{n/2, n/2}$ is
  immediately seen from this.

\end{proof}

\begin{theorem}
  \label{ch20theorem3proof2}
  \lean{mantel_amgm}
  \leanok
  Suppose $G$ is a graph on $n$ vertices without triangles. Then $G$
  has at most $\frac{n^2}{4}$ edges, and equality holds only when $n$ is even and $G$ is the
  complete bipartite graph $K_{n/2, n/2}$.
\end{theorem}
\begin{proof}
  \leanok
  The following proof of Theorem~3, using the inequality of the arithmetic and the geometric mean,
  is a folklore Book Proof. Let $\alpha$ be the size of a largest independent set $A$, and set $\beta = n - \alpha$.
  Since $G$ is triangle-free, the neighbors of a vertex $i$ form an independent set, and we infer
  $d_i \le \alpha$ for all $i$.

  The set $B = V \setminus A$ of size $\beta$ meets every edge of $G$.
  Counting the edges of $G$ according to their endvertices in $B$, we obtain
  $|E| \le \sum_{i \in B} d_i$.
  The inequality of the arithmetic and geometric mean now yields
  \[
  |E| \le \sum_{i \in B} d_i \le \alpha\beta \le \left(\frac{\alpha + \beta}{2}\right)^2 = \frac{n^2}{4},
  \]
  and again the case of equality is easily dealt with.
\end{proof}
